%   Filename    : acknowledgment.tex 
\begin{center}
	\textbf{Acknowledgment}
\end{center}

We would like to express our heartfelt gratitude to all those who have supported us throughout this research journey.

To our esteemed adviser, \textbf{Sir Dimzon}---your guidance and expertise have been invaluable, and we are deeply appreciative of the time, effort, and resources you have provided us. Your insights have significantly shaped our understanding of the subject matter.

To the beloved and renowned linguist, \textbf{Lolo David}---your unwavering dedication to the Akeanon language and your generosity in sharing your knowledge and resources have been truly inspiring. Your expertise has not only enriched our research but has also given us the confidence to explore new ideas and directions. We are deeply grateful for your encouragement and for the solid foundation you have helped us build. And of course, to \textbf{Ms. Hazel Cipriano}---your linguistic expertise and support have been instrumental in our research. Your contributions have greatly enhanced the quality of our work, and we are thankful for your unwavering support.

To \textbf{Dr. John Orbista and the College of Teacher Education of Aklan State University Banga campus}---your collaboration has been essential in ensuring the relevance and accuracy of our research. We are especially grateful for your assistance in providing us participants for this study.

To the people of \textbf{Brgy. Unidos} in Nabas, \textbf{Brgy. Motag} in Malay, \textbf{Brgy. Mayapay} in Buruanga, \textbf{Brgy. Guadalupe} in Libacao, and our \textbf{Aklanon friends and families}---your willingness to participate in our study has been invaluable. Your contributions have provided us with the necessary data to develop our Akeanon speech corpus, and we are grateful for your time and effort.

To \textbf{Dr. Anthea R. Redison} and \textbf{Dr. John E. Barrios}---your invaluable insights, knowledge, and constructive feedback have greatly refined our research and deepened our understanding of the Akeanon language. Your support in validating our data and guiding our research direction has significantly strengthened the credibility and relevance of our constructed text corpus to the Akeanon community. We are also grateful to Dr. Barrios for granting us permission to use his literary work as part of our text corpus.

To \textbf{Tita Linda Sarabia-Belayro}, \textbf{Mrs. Pamela Joy Mindanao}, \textbf{Sir Theodore Pastrana}, and the never-forgotten, \textbf{Sir Melchor Cichon}---your literary works served as the foundation for building our text corpus. We are deeply grateful for your contributions to Akeanon literature, which have greatly enriched our research. We especially honor the memory of Sir Melchor Cichon, whose recent passing is a great loss to the community; his legacy and dedication to the Akeanon language will always be remembered.

We would also like to thank YouVersion in granting us permission to use their resources from Bible.com, which have been instrumental in the development of both our text and speech corpus.

To our \textbf{families and friends}---your unwavering support, patience, and encouragement have been our pillars of strength throughout this journey. We are grateful for your understanding and belief in our abilities. To our \textbf{instructors and mentors} at the University of the Philippines Visayas, your teachings and guidance have laid the groundwork for our academic pursuits. We are thankful for the knowledge and skills you have imparted to us.

Finally, we would like to acknowledge the \textbf{Akeanon language} itself---a rich and vibrant language that has inspired us to delve deeper into its linguistic intricacies. We hope that our research will contribute to the preservation and appreciation of Akeanon for future generations.

As proud Aklanons, we are deeply honored to contribute to the preservation and promotion of our language and culture. The support and encouragement from fellow Aklanons have motivated us to pursue this research with passion and dedication.

Abo gid nga saeamat!