%   Filename    : dedication.tex 
\begin{center}
	\textbf{Dedication}
\end{center}

\textbf{From Jose:}

After ten years and enough coffee to caffeinate a small country, this Special Problem is finally done. I’m not sure if this is a computer science degree or a lifetime achievement award—but either way, here we are.

I would like to express my deepest gratitude to \textbf{Sir Francis Dimzon}, my SP Adviser, for their unwavering support, insightful guidance, and constant encouragement throughout this research journey. Their expertise and patience have been invaluable, and I am truly grateful for the opportunity to learn under their mentorship.

To the faculty of the \textbf{Computer Science Department}, thank you for teaching me not just how to code, but how to debug my life along the way. Your lectures, support, and endless deadline extensions have been greatly appreciated. I would also like to thank my groupmates, \textbf{Joshua} and \textbf{Mariefher}, for their constructive feedback and valuable suggestions that greatly improved the quality of this work.

To my komsai kids, \textbf{Rainer}, \textbf{Luis}, \textbf{Kurt}, \textbf{Jeff}, \textbf{Hamir}, \textbf{Chester}, and \textbf{Josiah}, I thank you for sticking with me despite our age gaps, it makes me proud being your upperclassmen. To \textbf{Ma'am Isay}, thank you for our weekly chats, it made the later parts of my college life more bearable. To all the \textbf{Dorm Staff} in the Dorms I stayed inside the campus, Thank you for providing me a home to where I can rest and concentrate. To my friends and classmates, thank you for the camaraderie and moral support it was fun meeting you all throughout my overdued stay.

To my family—especially my parents and my grandmother—for their unconditional love, endless patience, and belief in me. Your sacrifices and encouragement have been the foundation of all my achievements.

To everyone who helped, encouraged, tolerated, or simply waited: thank you. This Special Problem, and this long-overdue chapter, is dedicated to you.

\newpage


\textbf{From Joshua:}

First and foremost, to our dear \textbf{God}—thank You for the gift of life, the strength to endure, and the wisdom to pursue this path. Your divine guidance and grace have been my source of hope and courage throughout this journey.

This work is lovingly dedicated to the resilient and proud speakers of the \textbf{Akeanon language}—past, present, and future—whose voices, stories, and traditions have shaped the soul of Aklan.

To my beloved mother, \textbf{Glenda Villanueva}—your unwavering love, strength, and quiet encouragement have been the foundation of all my pursuits. This achievement is as much yours as it is mine.

To my sister, \textbf{Jhea Villanueva}—thank you for your patience, light, and steadfast support.

To my Tita and Titos, especially \textbf{Ante Nenia Alejandro} and \textbf{Tita Melma Villanueva}—your prayers and encouragement carried me through.

To my groupmates, \textbf{Mariefher Grace Villanueva} and \textbf{Nong Jose Fortaleza}—thank you for your dedication, teamwork, and perseverance.

To my friends—too many to name, but never forgotten—thank you for the laughter, encouragement, and quiet strength you offered along the way.

To \textbf{Sir Rex Aguirre}, thank you for your generosity and help with the Malaynon translation.

To the local communities, our linguist \textbf{Ms. Hazel Ann Cipriano}, our collaborators \textbf{Dr. John Orbista} and the \textbf{ASU Banga faculty and students}, our validators \textbf{Dr. Anthea R. Redison} and \textbf{Dr. John E. Barrios}, and to \textbf{Dr. R. David Zorc}, whose pioneering work in Akeanon linguistics laid the foundation for our study—thank you for your invaluable support, participation, and contributions.

To \textbf{Sir Francis D. Dimzon}, our adviser—thank you for your mentorship and unwavering support throughout this research.

And finally, to all who advocate for \textbf{Philippine languages} and \textbf{digital inclusion}—may this project serve as a meaningful contribution to the preservation and promotion of our linguistic and cultural heritage.

\newpage

\textbf{From Mariefher:}

First and foremost, I don't think words are enough to express my gratitude to all the people who have supported me throughout this journey. This work is a product of countless hours of hard work, dedication, and the unwavering support of those around me.

I would like to dedicate this work to my family, especially to my parents, \textbf{Marita} and \textbf{Porferio Villanueva}. I literally would not be here without your love, support, and sacrifices. Thank you for believing in me and for always being there for me. To \textbf{Mommy} and \textbf{Mama Virgie}, this is for you as well.

To my friends, especially \textbf{KC}, \textbf{Ava}, \textbf{Anaiah}, \textbf{Earl}, \textbf{Hanna}, \textbf{Louise}, \textbf{Cynel}, \textbf{Ivanne}, \textbf{Victor}, and too many to mention but you know who you are—your constant encouragement has always given me strength to continue. I am forever grateful to you all for always having my back, especially when I needed it the most. Shoutout also to my roommates throughout the years—\textbf{Shea}, \textbf{Abi}, \textbf{Cole}, \textbf{Dleamnor}, \textbf{Claire}, \textbf{Andy}, and \textbf{Learns}—thank you for the laughs, companionship, and for bearing with me, knowing my sleep schedule.

To \textbf{BINI}, indeed "buhay ay 'di karera". Thank you for being my inspiration and giving me more courage to move forward. You made me feel less lonely when I was alone.

To \textbf{Joshua} and \textbf{Jose}, thank you for your hard work, collaboration, and unwavering support throughout this journey. I am grateful for the shared challenges, laughter, and determination that made this whole thing possible.

To our adviser, \textbf{Sir Dimzon}, thank you for your invaluable guidance, patience, and encouragement throughout this journey. Your mentorship has been instrumental in shaping not only this work but also my growth as a student and individual. I am deeply grateful for your support and belief in our team.

To my instructors, especially \textbf{Ma'am Cala-or}, at the \textbf{University of the Philippines Visayas}, thank you for your guidance, patience, and dedication in teaching and inspiring me throughout my academic journey. Your encouragement and support have made a significant impact on my growth and learning.

And to the people who have been key contributors to this research, your expertise, time, and willingness to share your knowledge have been invaluable. Special thanks to \textbf{Lolo David} and \textbf{Hazel}, \textbf{Dr. Orbista}, \textbf{Sir John}, and the ever-supportive, \textbf{Ma'am Redison}. Your support and contributions have greatly enriched this work, and I am deeply grateful for your involvement.

As I close this chapter, I look back at the sleepless nights, the tears shed in silence, and the moments when giving up seemed easier than moving forward. Yet, because of your unwavering faith in me, I found the courage to persevere. This work is not just a testament to my efforts, but a tribute to the love, friendship, and support that carried me through the darkest and brightest days. To everyone who stood by me---this achievement is as much yours as it is mine. Thank you, from the depths of my heart.